Quantifying Stability and Change in Personal Culture Using Panel Data

### 2023-09-19 manuscript.Rmd
draft updates: case, results, measure

In this paper, we try to overcome previous debates and move away from the “tournament of models” approach toward a quantification of the processes that lead to differences in personal culture. Drawing on the seven panel data sets analyzed in these debates, we use established methods to quantify the amount of variance in responses that is due to between-people differences in baselines and the amount of variance in responses that is due to linear within-person change over time. This measure gives us a sense of the importance of intrapersonal change over the course of the panel relative to the differences that exist between people when they start. We also explore features of questions and populations that explain differences in within-person change over time.

The pattern of results help clarify and, hopefully, transcend previous debates. First, we observe similar patterns across the data sets that have produced conflicting interpretations. In all datasets we see similar rates of individual-level change over time across a broad range of measures of personal culture. But quantifying these processes also reveals that intrapersonal change plays a substantially smaller role in explaining the variance in peoples’ personal culture than between-person differences in their baselines. Whether this amount of intrapersonal change is “meaningful” ultimately remains a theoretical question reflected in prior expectations concerning the question, population, and time. Nevertheless, the results suggest that the intrapersonal change observed (in some cases over more than 10 years) does not seem to play a substantial role in explaining personal cultural differences between adults. While adding to existing debates, the quantification we propose extends them in new ways that enables scholars to more clearly adjudicate theoretical mechanisms at play. We provide a series of illustrative examples of the utility of this approach for addressing different theoretical questions.

@@

There is no reason to think that only one of these models is “correct” in all times and places. A population observed over some time frame contains a mix of people who are changing and people who are not (or some people making very small or imperceptible changes in culture and others making perceptible changes). Instead, different perspectives argue that each of these ideal-typical models is more operative at different times, for different people, and for different elements of personal culture. For example, adolescence and early adulthood, typically viewed  as a “formative period” for personal cultural development, is likely characterized by higher rates of “active updating,” while middle age and later life are potentially characterized more by “settled dispositions” [@alwin1991; @krosnick1989; @eaton2009]. Similarly, highly salient issues, such as views around gay rights in the 2010s; issues that see substantial elite realignments, such as views around the Vietnam conflict in the 1970s [@zaller1992]; and novel issues of public opinion, such as views around vaccines during the Covid-19 pandemic, might be characterized by active updating, while low-salience, established issues are characterized by stability.

@@

However, the claim that change is a relatively infrequent phenomenon has been difficult to square with research identifying individual change across a number of cultural dimensions, such as morality [@brocic2021], trust [@mewes2021], and concerns about immigration [@kratz2021]. Similarly, there are longitudinal studies showing that cues from political elites can change a person’s position on specific policy issues [@slothuus2021; @zaller1992] and that changes in a person’s close contacts and acquaintances can change their attitudes on group-related politics [@gelman2021; @dellaposta2018]. Given how often we observe people change their personal culture, it is hard to accept that adults do not change.

@@

In his tournament of models, Lersch finds broad support for the LCAM in data from five countries, with the LCAM favored over the AUM and SDM on 297 of 428 questions he analyzed and the SDM favored over the other two models on 112 items, with the remaining questions producing inconclusive results. He concludes that “new experiences over the life course, which expose individuals to new social environments, including novel interactions, and different sets of resources, can persistently move individuals’ personal culture in novel directions” [@lersch2023: p. 243-244].

However, despite differences in how these researchers interpret their findings, the empirical results are not far apart. Kiley and Vaisey [-@kiley2020] found that the majority of questions they explored favored the AUM, meaning they showed some evidence of durable change among adults, a finding broadly compatible with Lersch’s results. At the same time, while Lersch [-@lersch2023] found that the LCAM was preferred in the majority of cases he examined, durable change does not account for a large proportion of variance explained on these items. On average, people changed only about .07 standard deviations over 10 years. Again, this is consistent with Kiley and Vaisey’s assertion that durable change is often small. And a quarter of the items Lersch studied still preferred the SDM, suggesting that even with looser assumptions about the process of change, more cases, and more waves, many questions show no evidence of durable change. Relatedly, in most studies purporting to find evidence of durable change in peoples’ personal culture, the change is often quite small. For example, Brocic and Miles [-@brocic2021] find that completing graduate degrees in humanities, arts, and social sciences shift peoples’ moral relativism only about .2 standard deviations on average compared to people with no college degree, and this is the largest effect they identify. And studies of aggregate change show that, even on items where cohort effects dominate, some people do change over time [@vaisey2016]. In other words, despite different interpretations, the results of previous work are in many ways highly consonant.

@@

The shortcoming of current debates is that they focus on whether people *ever* change, a question that is not theoretically interesting since the answer is almost always “yes.”  As noted previously, if we observe enough people for long enough, we will likely observe some of them making durable changes. A lack of evidence for change in one question might simply indicate that the question was not asked at a clear enough level of resolution, for a long enough time frame, or to enough people to find evidence of change. Conversely, finding evidence of change tells us little about how much change is present in a population and how meaningful that change is for social differences.

@@

In this approach, there are two numbers that each reflect a combination of relevant theoretical processes. First, interpersonal differences at the start of the survey, what Lersch calls “early imprinting,” reflects the accumulated experiences of people prior to entering the panel survey that shape how they answer questions. While this is typically associated with a “formative period” or a “settled disposition” (hence the name early imprinting), it could also reflect transformative experiences that happen at any age, so long as they occurred before the start of the panel and leave a lasting effect on how people answer questions. For example, among people already retired upon entering the panel, their “early imprinting” might reflect this life course transition. To some extent, then, this number also reflects variation across people in relevant social roles or statuses at the start of the panel that are important in shaping dispositions.

The second, the amount of intrapersonal linear change over the course of the survey, what Lersch calls “persistent change” or “adaption” and Kiley and Vaisey call “active updating,” reflects durable changes in personal culture over time. Lersch suggests that these changes are brought on by social triggers such as moving into a new environment, adopting new social roles, or broader changes in the social structure.

@@

We start with Lersch’s [-@lersch2023] Life Course Adaption Model, which formalizes survey responses at time $t$ as a function of individual-level random intercepts and slopes for survey age, what is often called a mixed-effects growth curve model. This model formally assumes a set of propositions about change that reflect the theoretical debate to this point. First, consistent with the settled disposition model, it assumes that people start the survey with cultural differences, modeled as random intercepts for each respondent. Second, it assumes that people change over time, taking the form of random slopes for each respondent as a function of age. Third, it assumes that people deviate around this baseline randomly with age, reflecting “fluctuation” or short-term non-persistent change. Finally, it assumes that people draw on earlier time points in reacting to new events. Formally, this can be written as

@@

As a first step in our analysis, we fit LCAMs to each of the 609 measures of personal culture outlined above. We then use these models to derive two measures of residual variance that correspond to our theoretical processes. First, to measure the importance of interpersonal differences, we calculate the residuals derived from measuring each respondent at the midpoint of their age trajectory, denoted as $\tau$. By dividing this value by the total variance in responses and subtracting it from 1, we calculate the proportion of variance explained by interpersonal differences. Second, we calculate residuals for additionally accounting for the age trajectory, denoted as $\delta$. Subtracting the variance explained by interpersonal differences and dividing this new value by the total variance give us the additional proportion of variance explained by intrapersonal change. These two components reflect total “explained” variance of the model. Our principal outcome of interest is the proportion of explained variance accounted for by intrapersonal change, which we calculate as

@@

@@

Our results proceed in three parts. We first explore the overall pattern of variance explained by interpersonal differences at baseline and intrapersonal change over time across the seven panels. Next, we analyze the association between the relative proportion of variance explained by intrapersonal change as a function of question, panel, and sample features. Finally, we highlight a few specific comparisons to illustrate the value of quantification relative to adjudication for further exploring theoretical questions.

@@

Figure 1 plots the proportion of variance explained by interpersonal differences (“between”) and intrapersonal change (“within”) for each of the 609 questions analyzed, plotted separately by panel. We highlight two top-line findings. First, despite the differences in interpretation present in previous works, results from the GSS appear remarkably consistent with results from the other panels. All panels show a wide range of the proportion of explained variance attributable to intrapersonal change, with all panels including questions that show essentially no variance explained by intrapersonal change and all panels including a few questions where more than 20 percent of explained variance is attributable to intrapersonal change. Across all the questions, interpersonal differences explain a much larger share of the variance in responses. Again, this is to be expected. Interpersonal differences capture not just pre-adult socialization, but all accumulated experiences up to the start of the panel.

The substantive importance of intrapersonal change depends on a range of factors, including how long the panel runs, whether assumptions about linear change hold, and whether the panel is capturing a distinctly turbulent period or a distinctly stable period for the item under question. However, we observe an average of about 9.7 percent of explained variance being attributed to intrapersonal change across all the questions. If we assume that the period under observation is “typical” for a question –-not a time of extremely heightened (or lowered) sensitivity or change-– then it does not seem realistic for this process of intrapersonal change to explain a large share of interpersonal differences.

@@

Figure 2 plots the results from an ordinary least squares regression of proportion of explained variance attributable to intrapersonal change as a function of question, panel, and sample features. We find that the more response options respondents are given and the larger the sample, the larger the proportion of explained variance attributable to intrapersonal change. We interpret these coefficients as suggesting that greater resolution on a question makes it easier to detect and model change. The earlier a question was asked, even net of how long it has been asked, is associated with lower proportion of intrapersonal change.

The most notable result in the meta analysis is the negative coefficient attached to the span of years covered by the question. Theories of personal cultural change that link changes in personal culture to social experiences, including the LCAM, suggest that the longer we observe respondents, the more likely people are to undergo such experiences and therefore the more variance would be explained by intrapersonal change. The fact that we find a negative coefficient attached to this variable –-the longer we observe people, the less variance is explained by intrapersonal change-– is a challenge to that assumption.

As a further test of this coefficient, we performed a within-question analysis, comparing the proportion of variance explained by intrapersonal change by the same question when using the full panel as when we dropped the final wave, and therefore reduced the total duration of observation for the question. Results from this analysis are presented in Appendix B, which contradict the coefficient from the regression model. The longer we observe a question, the higher proportion of explained variance is attributable to intrapersonal change. We interpret this combination of findings as suggesting that the kinds of questions asked for a longer time period tend to demonstrate less intrapersonal change than questions asked over shorter periods.

@@

@@

This paper set out to clarify previous findings about the prevalence of durable change in personal culture -- the attitudes, beliefs, values, and preferences that manifest at the individual level. Relying on an approach that quantified the amount of response variance attributable to either intrapersonal change or interpersonal differences at baseline, we found a consistent pattern across data sets previously used in this debate. We see evidence of intrapersonal change over time, with some measures of personal culture displaying quite high levels of intrapersonal change relative to the amount of variance between people. On some questions (such as factors that affect one’s life satisfaction or attitudes about government benefits), it is highly plausible that changes in personal culture as a result of peoples’ experiences as adults are the primary explanation for differences in personal culture.

However, the amount of intrapersonal change is often small (approximately 10% across all surveys) relative to interpersonal differences. On many questions, the amount of explained variance attributable to intrapersonal change is essentially zero. These include questions about civil liberties, abortion, interindividual trust, or sense of civic duty. On these questions, assuming the window of observation is not uniquely stable, the amount of variance attributable to intrapersonal change does not seem sufficient to explain how and why adults end up with different personal cultures. For these items, the explanation for personal cultural differences seems to reside principally in early-life experiences.

It is difficult to draw a single conclusion about the relative importance of intrapersonal change in explaining cultural differences across a wide range of questions measured on different scales and over different durations. A key assertion of this work is that blanket statements of “change” versus “stability” as the single pattern among adults is likely to be insufficient. Change and stability are always present among adults, and no viable theoretical model expects just one. But the overall picture, with important caveats, suggests that intrapersonal change during adulthood is not the dominant explanation for cultural differences among adults for most items we measure in surveys. There simply is not enough evidence of intrapersonal change on items relative to the amount of difference we observe across people.

To be clear, this does not mean that adults do not change or are incapable of changing. On basically all questions, some members of the sample change their personal culture. And our results should not be taken to imply that intrapersonal change is not important or consequential. Small numbers of changers can have important consequences in the aggregate. A shift from 49 percent supporting gay marriage to 51 percent supporting it can have significant political ramifications, even as most respondents remain fixed in their preferences.

>>>>

In this paper, we try to overcome previous debates and move away from the “tournament of models” approach toward a quantification of the processes that lead to differences in personal culture. Drawing on the seven panel data sets analyzed in these debates, we use established methods to quantify the amount of variance in responses that is due to between-people differences in baselines and the amount of variance in responses that is due to linear within-person change over time. This measure gives us a sense of the importance of intrapersonal change over the course of the panel relative to the differences that exist between people when they start, a quantity with more diverse applications than a simple judgment about whether change happened. We also explore features of questions and populations that explain differences in within-person change over time.

The pattern of results help clarify and, hopefully, transcend previous debates. We observe similar patterns across the data sets that have produced conflicting interpretations. In all datasets we see similar rates of individual-level change over time across a broad range of measures of personal culture. But quantifying these processes also reveals that intrapersonal change plays a substantially smaller role in explaining the variance in peoples’ personal culture than between-person differences in their baselines. Whether this amount of intrapersonal change is “meaningful” ultimately remains a theoretical question reflected in prior expectations concerning the question, population, and time. Nevertheless, the results suggest that the intrapersonal change observed (in some cases over more than 10 years) does not seem to play a substantial role in explaining personal cultural differences between adults.

While adding to existing debates, the quantification we propose extends them in new ways that enables scholars to more clearly adjudicate theoretical mechanisms at play. We provide an example by exploring differences in the relative importance of intrapersonal change for people with and without college degrees, comparing both within and across questions in ways that the tournament of models cannot.

@@

There is no reason to think that only one of these models is “correct” in all times and places. A population observed over some time frame contains a mix of people who are changing and people who are not (or some people making very small or imperceptible changes in culture and others making perceptible changes). Instead, different perspectives argue that each of these ideal-typical models is more operative at different times, for different people, and for different elements of personal culture. For example, adolescence and early adulthood, typically viewed  as a “formative period” for personal cultural development, is likely characterized by higher rates of active updating, while middle age and later life are potentially characterized more by settled dispositions [@alwin1991; @krosnick1989; @eaton2009]. Similarly, highly salient issues, such as views around gay rights in the 2010s; issues that see substantial elite realignments, such as views around the Vietnam conflict in the 1970s [@zaller1992]; and novel issues of public opinion, such as views around vaccines during the Covid-19 pandemic, might be characterized by active updating, while low-salience, established issues are characterized by stability.

@@

However, the claim that change is a relatively infrequent phenomenon has been difficult to square with research identifying durable individual change across a number of cultural dimensions, such as morality [@brocic2021], trust [@mewes2021], and concerns about immigration [@kratz2021], as a result of social experiences. Similarly, there are longitudinal studies showing that cues from political elites can change a person’s position on specific policy issues [@slothuus2021; @zaller1992] and that changes in a person’s close contacts and acquaintances can change their attitudes on group-related politics [@gelman2021; @dellaposta2018]. Given how often we observe people change their personal culture, it is hard to accept that adults do not change.

@@

In his tournament of models, Lersch finds broad support for the LCAM in data from five countries, with the LCAM favored over the AUM and SDM on 297 of 428 questions he analyzed and the SDM favored over the other two models on 112 items, with the remaining questions producing inconclusive results (no questions preferred the AUM). He concludes that “new experiences over the life course, which expose individuals to new social environments, including novel interactions, and different sets of resources, can persistently move individuals’ personal culture in novel directions” [@lersch2023: p. 243-244].

Despite differences in how these researchers interpret their findings, the empirical results are not far apart. Kiley and Vaisey [-@kiley2020] found that the majority of questions they explored favored the AUM, meaning they showed some evidence of durable change among adults, a finding broadly compatible with Lersch’s results. At the same time, while Lersch [-@lersch2023] found that the LCAM was preferred in the majority of cases he examined, durable change does not account for a large proportion of variance explained on these items. On average, people changed only about .07 standard deviations over 10 years. Again, this is consistent with Kiley and Vaisey’s assertion that durable change is often small. And a quarter of the items Lersch studied still preferred the SDM, suggesting that even with looser assumptions about the process of change, more cases, and more waves, many questions show no evidence of durable change. Relatedly, in most studies purporting to find evidence of durable change in peoples’ personal culture, the change is often quite small. For example, Brocic and Miles [-@brocic2021] find that completing graduate degrees in humanities, arts, and social sciences shift peoples’ moral relativism only about .2 standard deviations on average compared to people with no college degree, and this is the largest effect they identify. And studies of aggregate change show that, even on items where cohort effects dominate, there is always evidence that some people change over time [@vaisey2016]. In other words, despite different interpretations, the results of previous work are in many ways highly consonant.

@@

The shortcoming of current debates is that they focus on whether people *ever* change, a question that is not theoretically interesting since the answer is almost always “yes.”  As noted previously, if researchers observe enough people for long enough, they will likely observe some of them making durable changes. A lack of evidence for change in one question might simply indicate that the question was not asked at a clear enough level of resolution, for a long enough time frame, or to enough people to find evidence of change. Conversely, finding evidence of change tells us little about how much change is present in a population and how meaningful that change is for social differences.

@@

In this approach, there are two numbers that each reflect a combination of relevant theoretical processes. First, interpersonal differences at the start of the survey, what Lersch calls “early imprinting,” reflects the accumulated experiences of people prior to entering the panel survey that shape how they answer questions. While this is typically associated with a formative period or a settled disposition (hence the name early imprinting), it could also reflect transformative experiences that happen at any age, so long as they occurred before the start of the panel and leave a lasting effect on how people answer questions. For example, among people already retired upon entering the panel, their “early imprinting” might reflect this life course transition. To some extent, then, this number also reflects variation across people in relevant social roles or statuses at the start of the panel that are important in shaping dispositions.

The second set of processes, what Lersch calls “persistent change” or “adaption” and Kiley and Vaisey call “active updating,” reflects durable changes in personal culture over time and is captured by the amount of intrapersonal linear change over the course of the survey. Lersch suggests that these changes are brought on by social triggers such as moving into a new environment or adopting new social roles. They could also reflect the diffusion of new culture across social networks, changes in social network composition (or knowledge about who is in a person's network), new cues from political elites or other cultural leaders, the emergence of new issues in politics or culture, or other large-scale social shifts.

@@

We start with Lersch’s [-@lersch2023] Life Course Adaption Model, which formalizes survey responses at time $t$ as a function of individual-level random intercepts and slopes for survey age, what is often called a mixed-effects growth curve model. This model formally assumes a set of propositions about change that reflect the theoretical debate to this point. First, consistent with the settled disposition model, it assumes that people start the survey with cultural differences, modeled as random intercepts for each respondent. Second, it assumes that people change over time, taking the form of random slopes for each respondent as a linear function of time. Third, it assumes that people deviate around this baseline randomly over time, reflecting “fluctuation” or short-term non-persistent change. Finally, it assumes that people draw on earlier time points in reacting to new events. Formally, this can be written as

@@

As a first step in our analysis, we fit LCAMs to each of the 609 measures of personal culture outlined above. We then use these models to derive two measures of systematic variance that correspond to the two collections of theoretical processes outlined above: interpersonal differences at baseline and intrapersonal change over time. First, to measure the importance of interpersonal differences, we calculate the sum of the squared residuals derived from measuring each respondent at the midpoint of their age trajectory.[^midpoint] We then divide this value by the total sum of squares and subtract it from from 1, producing a value we denote as $\tau$, or the proportion of variance explained by interpersonal differences. This number is equivalent to an $R^2$ value generated by using random intercepts for all respondents. Second, we calculate the sum of squared residuals generated by additionally accounting for the age trajectory. We divide this number by the total sum of squares and subtract it from 1. We then subtract $\tau$ from this value. This gives us a value we denote as $\delta$, the additional proportion of variance explained by intrapersonal change.

These two numbers reflect the total systematic or "explained" variance of the model that can be attributed to either differences across people or linear change over the course of the panel.[^explained] Our principal outcome of interest is the proportion of systematic variance accounted for by intrapersonal change, which we calculate as

@@

[^midpoint]: We use the scale midpoint, rather than the actual difference at baseline, to better capture the differences between people independent of their age trajectories.

[^explained]: This does not mean that other factors might not explain additional variance if they were included in the model. For example, peoples' responses about religious beliefs and participation might systematically vary by the day of the week or month in which they are surveyed. This variation is moved into the "random" or non-systematic component of the model for us.

We can think of this number as akin to an Intraclass Correlation Coefficient focused only on the systematic components of the model while ignoring residual variance that might be attributable to measurement error or non-persisting change [see also the variance decomposition of period and cohort by @vaisey2016]. The higher the number is, the more variance in responses is attributable to intrapersonal change during the panel. The lower the number, the less intrapersonal change seems to play a large role in explaining cultural differences.

Because this quotient summarizes variance and therefore has no natural referent, it is hard to say at which values we observe "a lot" of intrapersonal change. But summarizing questions in this way allows us to compare the relative importance of our two broad processes across a range of questions that might have very different levels of non-systematic variance, giving us a baseline sense of the relative prevalence of these two processes for each question (something not achievable under the tournament of models approach), and allowing us to compare which kinds of questions show more or less intrapersonal change relative to interpersonal differences. While the tournament of models approach allowed researchers to say whether questions in general showed evidence of updating, this approach allows us to make more specific claims about the prevalence of these theoretical processes for particular questions relative to each other.

@@

Our results proceed in three parts. We first explore the overall pattern of systematic variance explained by interpersonal differences at baseline and intrapersonal change over time across the seven panels. Next, we analyze the association between the relative proportion of systematic variance explained by intrapersonal change as a function of question, panel, and sample features. Finally, we highlight a few specific comparisons to illustrate the value of quantification relative to adjudication for further exploring theoretical questions.

@@

Figure 1 plots the proportion of systematic variance attributable to interpersonal differences (“between”) and intrapersonal change (“within”) for each of the 609 questions analyzed, plotted separately by panel. All panels show a similar range of the proportion of systematic variance attributable to intrapersonal change, with all panels including questions that show essentially no systematic variance attributable to intrapersonal change and all panels including a few questions where more than 20 percent of systematic variance is attributable to intrapersonal change. Across all questions, interpersonal differences explain a much larger share of the systematic variance in responses. Again, this is to be expected. Interpersonal differences capture not just pre-adult socialization, but all accumulated experiences up to the start of the panel that might have a bearing on personal culture. Despite the differences in interpretation present in previous works, results from the GSS are highly consistent with results from the other panels.

PARAGRAPH HERE ABOUT RESULTS: MOST/LEAST SYSTEMATIC VARIANCE. DISTRIBUTION. NEED THAT INFO.

Appendix A presents the estimated proportion of variance attributable to interpersonal differences and intrapersonal change, as well as the proportion of residual variance, for each question analyzed. In general, interpersonal differences are almost always the largest component of the total variance and tend to explain between 50 and 70 percent of total variance, while intrapersonal change is always the smallest, typically explaining between 0 and 15 percent of total variance. Residual variance tends to account for between 20 to 40 percent of variance (though on several questions residual variance is greater than 50 percent).

The substantive importance of intrapersonal change for understanding cultural differences depends on a range of factors, including how long the panel runs, whether assumptions about linear change hold, and whether the panel is capturing a distinctly turbulent period or a distinctly stable period for the item under question. However, we observe a mean of about 9.7 percent of explained variance being attributed to intrapersonal change across all the questions. If we assume that the period under observation is “typical” for a question –-not a time of extremely heightened (or lowered) sensitivity or change-– then it does not seem realistic for this process of intrapersonal change to explain a large share of interpersonal differences.

@@

Figure 2 plots the results from an ordinary least squares regression of proportion of systematic variance attributable to intrapersonal change as a function of question, panel, and sample features. We find that the more response options respondents are given and the larger the sample, the larger the proportion of systematic variance is attributable to intrapersonal change. We interpret these coefficients as suggesting greater resolution on a question makes it easier to detect and model change. The earlier a question was asked, even net of how long it has been asked, the lower the proportion of systematic variance is attributable to intrapersonal change.

The most notable result in the meta analysis is the negative coefficient attached to the span of years covered by the question. Theories of personal cultural change that link changes in personal culture to social experiences, including the LCAM, suggest that the longer we observe respondents, the more likely people are to undergo potentially transformative experiences and therefore the more variance would be explained by intrapersonal change. The fact that we find a negative coefficient attached to this variable –-the longer we observe people, the less variance is explained by intrapersonal change-– is a challenge to that assumption.

As a further test of this coefficient, we compared the proportion of systematic variance attributable to intrapersonal change when using the full duration of a panel compared to when when we dropped the final wave and therefore reduced the total duration of observation for the question. If the coefficient reflects a true negative "causal" effect of duration on systematic variance attributable to intrapersonal change, we should see that same effect within questions. Results from this analysis are presented in Appendix B and contradict the coefficient from the regression model. The longer we observe a question, the higher proportion of systematic variance is attributable to intrapersonal change. We interpret this combination of findings as suggesting that the kinds of questions asked for a longer time period tend to demonstrate less intrapersonal change than questions asked over shorter periods, rather than a true function of time.

@@

While these broad results help clarify previous debates, they mostly reframe previous empirical results. To better illustrate the value of our approach, we turn to a specific empirical example: the relative importance of intrapersonal change and interpersonal difference for explaining cultural differences by education levels.

Previous work has established a positive relationship between education and attitude stability, especially on issues related to American politics. This stability is often attributed to education facilitating "chronic knowledge" -- a general understanding of the domain of American politics, including the positions held by major parties and political figures and how issues relate to one another at a logical or socio-logical level [@zaller1992; @alvarez2002; @boutyline2017]. Because they have more knowledge of American politics, people with a college education are better able to consistently connect the considerations in their cognition with the answer choices they are presented with in a survey.

This work has tended to focus on the fact that college-educated Americans give responses that are less likely to be affected by measurement error or short-term influences [@zaller1992; @alwin2003]. However, this focus on the difference in the non-systematic (residual) component of variance across groups obscures the fact that the relative importance of different components of systematic variance might differ across these two groups as well.

It could be the case that education, chronic knowledge, and attention to the political sphere makes intrapersonal change more important in explaining differences among people with higher levels of education. For example, these people might be more likely to make systematic changes in response to the emergence of new information, new issues, or political realignments, while those without chronic knowledge display more variance around their baselines without their baselines changing much [@zaller1992]. In contrast, it could be the case that the stability demonstrated by educated respondents reflects the fact that they have already formed durable opinions and are relatively closed off to new information. College could serve as a formative experience that solidifies attitudes for people who attend. But among people who do not attend college, later life experiences might prove more important in forming or changing attitudes, as these experiences potentially provide information that college-educated peers received earlier.

To test these competing propositions, we use the models presented above to calculate the total systematic variance for people with at least a bachelor's degree and people with less than a bachelor's degree at wave 1[^wave] of the General Social Survey's panels.[^missing] We focus on the GSS because it contains the largest number of questions tapping general political dispositions, which is the domain where education has proven particularly relevant for understanding attitude stability. The GSS also covers a particularly turbulent window of American politics from 2006 to 2014. This window covers the start of the Great Recession, debates about federal intervention in and regulation of Wall Street, the election of Barack Obama as the first black president in U.S. history, debates about the role of the federal government in the health care sector, the emergence of the Tea Party, and political realignment and clarification on the issue of gay marriage, among other topics.

[^wave]: A small number of respondents report different highest degrees at wave 1, wave 2, and wave 3. Some of this is due to measurement error, and some of it is due to a small number of people obtaining a higher degree during the four years of the panel. Estimating the panel with highest at wave 3 or highest degree reported across the panel produces functionally identical results.

[^missing]: We estimate the models presented in [earlier figure] using the full sample, but only use respondents who have information on their highest degree to generate the results presented in figures [these figures]. In total, ### respondents (%%% percent) reported no degree information.

Figure XXX plots the distribution of differences between people with less than a bachelor's degree and people with at least a bachelor's degree at wave 1 of the panel. Values greater than 0 would indicate that intrapersonal change is more important in explaining the responses of college graduates than those of people without a college degree, while values less than 0 would indicate that intrapersonal change is less important for explaining variation among people without a college degree than those with a college degree.

There is a clear pattern in [figure above]: for more than 80 percent of GSS items explored here, intrapersonal change is a larger component of systematic variance for people without a college degree than for people with a college degree (they have negative values in the figure). While most of these differences are small in absolute terms (less than 2 percentage points), several are greater than 5 percentage points. Given the distribution observed in Figure [the earlier one] showing that the systematic variance attributable to intrapersonal change averages [average across all questions], a 5 percentage point difference between groups is quite substantial.

To more clearly illustrate some of these differences, we highlight eight questions designed to tap general political dispositions: partisan identification (Democrat vs. Republican) and ideological identification (liberal vs. conservative) on seven-point scales; four questions about the government's role in improving the condition of the poor, paying people's medical bills, giving special treatment to blacks, and doing things that private businesses could do, measured on five-point scales; a question about whether the government should do more to reduce income differences, measured on a seven-point scale; and one question about whether blacks should be given preferences in hiring, measured on a five-point scale. Estimates for the proportion of systematic variance attributable to intrapersonal change on these eight questions, for both education groups, are presented in Figure XXX.

On all eight questions presented in Figure XXX, intrapersonal change explains less systematic variance for people with a college degree than for people without a college degree. There are also large differences across questions on the amount of systematic variance attributable to intrapersonal change for both groups. For example, on the question of whether the government should try to solve more problems or leave those problems to be solved by private businesses ("government do more or less"), less than one percent of the systematic variance is attributable to intrapersonal change for both groups. In other words, while people might vacillate on this question at random (37 percent of variance is residual for this question), there is functionally no evidence that people make systematic changes of opinion on this issue during the GSS panel.

In contrast, partisan identification and political ideology both show a larger proportion of systematic variance attributable to intrapersonal change, with quite large differences between people with a college degree and those without. Compared to the other questions explored here, change as adults (and, potentially, their experiences as adults) plays a much larger role in explaining why people report the partisan identification and political ideology they do. And this is particularly true for respondents who do not have a college degree. The amount of systematic variance in ideological identification that is attributable to intrapersonal change among non-college educated respondents is almost four times that of college-educated respondents.

We note here that this meaningful difference in the proportion of systematic variance attributable to intrapersonal change across education groups and across questions would not be detectable using previous methods. For partisan identification and political ideology, both college-educated respondents and people with less than college degree would likely favor the AUM or LCAM over the SDM because they show evidence of some members of the population making intrapersonal change. In other words, using the tournament of models approach obscures the fact that the relative importance of intrapersonal change in explaining differences on these two items is very different across these two groups, that intrapersonal change explains more systematic variance for ideological identification than partisan identification for people without a college degree, or that there appears to be more durable change on questions of affirmative action than on questions of government aid to black Americans.

It is important to note here that these differences should not be interpreted as causal effects of attending college on the importance of social experiences for cultural change. These results are potentially confounded by age, social class, race, gender, and other factors that explain selection into higher education. We simply use these estimates as a starting point for a broader discussion.

@@

This paper set out to clarify previous findings about the prevalence of durable change in personal culture -- the attitudes, beliefs, values, and preferences that manifest at the individual level. Relying on an approach that quantified the amount of systematic variance attributable to either intrapersonal change or interpersonal differences at baseline, we found a consistent pattern across data sets previously used in this debate. We see evidence of intrapersonal change over time on almost all questions, with some measures of personal culture displaying quite high levels of intrapersonal change relative to the amount of variance between people. On some questions (such as factors that affect one’s life satisfaction or attitudes about government benefits), it is highly plausible that changes in personal culture as a result of peoples’ experiences as adults are the primary explanation for differences in personal culture.

However, the amount of intrapersonal change is often small relative to interpersonal differences (approximately 10 percent across all surveys). On many questions, the amount of systematic variance attributable to intrapersonal change is essentially zero. These include questions about civil liberties, abortion, interindividual (kk note: generalized?) trust, or sense of civic duty. On these questions, assuming the window of observation is not uniquely stable, the amount of variance attributable to intrapersonal change does not seem sufficient to explain how and why adults end up with different personal cultures. For these items, the explanation for personal cultural differences seems to reside principally in early-life experiences.

It is difficult to draw a single conclusion about the relative importance of intrapersonal change in explaining cultural differences across a wide range of questions measured on different scales and over different durations of time. A key assertion of this work is that blanket statements of “change” versus “stability” as the single pattern among adults is likely to be insufficient. Change and stability are always present among adults, and no viable theoretical model expects just one process to explain all difference. But the overall picture, with important caveats, suggests that intrapersonal change during adulthood is not the dominant explanation for cultural differences among adults for most items we measure in surveys. There simply is not enough evidence of intrapersonal change on these items relative to the amount of difference we observe across people.

To be clear, this does not mean that adults do not change or are incapable of changing. On basically all questions, some members of the sample change their personal culture. And our results should not be taken to imply that intrapersonal change is not important or consequential. Small numbers of changers can have important consequences in the aggregate. A shift from 49 percent supporting gay marriage to 51 percent supporting it can have significant political ramifications, even as most respondents remain fixed in their preferences. Our results could also be interpreted as suggesting that opportunities for durable change or the social conditions that facilitate durable change are just extremely uncommon. Changes in social structures, the life course, or cultural expectations could make intrapersonal change a much more consequential explanation for cultural differences.

Furthermore, the results we present related to education and political views suggest that the relative importance of different theoretical processes might differ by group. In line with life course perspectives, the results presented here suggest that importance of different experiences for attitude change are potentially contingent on previous experiences. For example, to the extent that we want to explain why people have different personal culture, the answer is likely different for people who have a college degree and those who do not. Mid-life experiences appear to have more significant implications for the personal culture for people without a college degree than for people with a college degree.
